Estimating the size of a population through repeated sampling: a new view on capture-recapture procedures

### 2022-03-09 _900_output/paper/paper.tex
Finish the draft of paper body

\section{Model and likelihood}
Our starting point is a basic capture-recapture model with equal capture probabilities. Consider a closed population $\mathcal{P} = \{1,\ldots,N\}$ from which we independently draw $T > 1$ samples of size $n$ using some sampling scheme. Define $S_t$ to be the set of individuals included into sample $t$ and $\pi_i = \text{pr}(i \in S_t)$ to be the inclusion probability of $i$. Let $X_{i(T)}$ denote the frequency count of $i$ being included into $T$ samples
\begin{equation*}
X_{i(T)} = \sum_{t=1}^T \mathds{1}\{i \in S_t\};
\end{equation*}
If $\pi_i = n / N$ for all $i$, then $X_{1(T)}, \ldots, X_{N(T)}$ are modeled as identically and independently distributed binomial variables
\begin{equation*}
\text{pr}(X_{i(T)} = x_i; T, N, n) = \binom{T}{x_i} \left(\frac{n}{N}\right)^{x_i} \left(1 - \frac{n}{N}\right)^{T - x_i}
\end{equation*}
However, since we only observe $X_{i(T)} > 0$, the above distribution must be truncated at zero. For convenience purposes, it is also useful to partition $\mathcal{P}$ into $D = \cup_{t=1}^T S_t = \{i \in \mathcal{P}: X_{i(T)} > 0\}$ and $U = \mathcal{P} \setminus D = \{i \in \mathcal{P}: X_{i(T)} = 0\}$. Accordingly, population size can be expressed as a sum of cardinalities of $D$ and $U$, i.e. $N = N_D + N_U$, where the first term is known and the latter term is to be estimated.  The resulting likelihood of the data is
\begin{equation*}
\mathcal{L}(N_U) = \prod_{i \in D} \frac{\text{pr}(X_{i(T)} = x_i; T, N, n)}{1 - \text{pr}(X_{i(T)} = 0; T, N, n)} = \prod_{i \in D} \binom{T}{x_i} \frac{n^{x_i} (N_D + N_U - n)^{T-x_i}}{(N_D + N_U)^{T} - (N_D + N_U - n)^T} \\
\end{equation*}
Setting $T = 2$ and recognizing that $\sum_{i \in D} x_i = nT$ yields the following maximum likelihood estimator of $N_U$
\begin{equation*}
\widehat{N}_U = \frac{(n - N_D)^2}{2n - N_D}
\end{equation*}
Thus, the population size is estimated by
\begin{equation} \label{eq:1}
\widehat{N} = N_D + \widehat{N}_U = \frac{n^2}{2n - N_D}
\end{equation}
Equation \eqref{eq:1} is the special case of the Lincoln-Peterson estimator with fixed $n$ for each sample draw \citep{Pollock:1990}.

Assume now that $\pi_i$ varies across population units meaning that \eqref{eq:1} is no longer applicable. Let us consider the problem from the Bayesian perspective and assume a beta prior for inclusion probabilities with hyperparameters $\alpha$ and $\beta$
\begin{equation*}
f_{\pi_i}(\upsilon; \alpha, \beta) = \frac{\upsilon^{\alpha - 1}(1 - \upsilon)^{\beta - 1}}{\mathrm{B}(\alpha, \beta)}, \quad \mathrm{B}(\alpha, \beta) = \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha + \beta)};
\end{equation*}
where $\Gamma(x)$ is the gamma function. From Bayes' theorem, we have
\begin{equation*}
f_{\pi_i \mid X_{i(T)}} (\upsilon \mid x_i) = \frac{\text{pr}(X_{i(T)} = x_i \mid \upsilon; T) f_{\pi_i}(\upsilon; \alpha, \beta)}{\text{pr}(X_{i(T)} = x_i)}
\end{equation*}
The marginal likelihood can be found in a straightforward manner
\begin{equation*}
\text{pr}(X_{i(T)} = x_i) = \int_0^1 \text{pr}(X_{i(T)} = x_i | \upsilon)f_{\pi_i}(\upsilon)d\upsilon = \binom{T}{x_i} \frac{\mathrm{B}(\alpha + x_i, \beta + T - x_i)}{\mathrm{B}(\alpha, \beta)}
\end{equation*}
Conditioning on $X_{i(T)} > 0$ results in the following truncated distribution
\begin{equation*}
\text{pr}(X_{i(T)} = x_i \mid X_{i(T)} > 0; \alpha, \beta, T) = \frac{\text{pr}(X_{i(T)} = x_i)}{1 - \text{pr}(X_{i(T)} = 0)} = \binom{T}{x_i} \frac{\mathrm{B}(\alpha + x_i, \beta + T - x_i)}{\mathrm{B}(\alpha, \beta) - \mathrm{B}(\alpha, \beta + T)}
\end{equation*}
Following the empirical Bayes approach, we estimate hyperparameters $\alpha$ and $\beta$ by maximizing the marginal likelihood.
\begin{align*}
    \mathcal{L}(\alpha, \beta) &= \prod_{i \in D} \binom{T}{x_i} \frac{\mathrm{B}(\alpha + x_i, \beta + T - x_i)}{\mathrm{B}(\alpha, \beta) - \mathrm{B}(\alpha, \beta + T)} \\
    &= \prod_{i \in D} \binom{T}{x_i} \frac{\Gamma(\alpha + x_i)\Gamma(\beta + T - x_i)\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)\Gamma(\alpha + \beta + T) - \Gamma(\alpha)\Gamma(\beta + T)\Gamma(\alpha + \beta)} \\
\end{align*}
Using the recursive relation of the gamma function $\Gamma(x) = (x - 1)\Gamma(x - 1)$ and the fact that $T$ and $x_i$ are natural numbers, we can re-write the marginal likelihood as
\begin{align} \label{eq:2}
    \mathcal{L}(\alpha, \beta) &= \prod_{i \in D} \binom{T}{x_i} \frac{\Gamma(\alpha)\Gamma(\beta)\Gamma(\alpha + \beta) \prod_{j=1}^{x_i} (\alpha + x_i - j)\prod_{j=1}^{T - x_i} (\beta + T - x_i - j)}{\Gamma(\alpha)\Gamma(\beta)\Gamma(\alpha + \beta)\{\prod_{j=1}^T(\alpha + \beta + T - j) - \prod_{j=1}^T (\beta + T - j)\}} \nonumber \\
    &= \prod_{i \in D} \binom{T}{x_i} \frac{\prod_{j=1}^{x_i} (\alpha + x_i - j)\prod_{j=1}^{T - x_i} (\beta + T - x_i - j)}{\prod_{j=1}^T(\alpha + \beta + T - j) - \prod_{j=1}^T (\beta + T - j)}
\end{align}
\begin{thebibliography}{1}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

>>>>

\section{Model and likelihood}
Our starting point is a basic capture-recapture model with equal capture probabilities. Consider a closed population $\mathcal{P} = \{1,\ldots,N\}$ from which we independently draw $T > 1$ samples of size $n$ using some sampling scheme. Define $S_t$ to be the set of individuals included into sample $t$ and $\pi_i = \text{pr}(i \in S_t)$ to be the inclusion probability of $i$. Let $X_{i(T)}$ denote the frequency count of $i$ being included into $T$ samples
\begin{equation*}
X_{i(T)} = \sum_{t=1}^T \mathds{1}\{i \in S_t\};
\end{equation*}
If $\pi_i = n / N$ for all $i$, then $X_{1(T)}, \ldots, X_{N(T)}$ are modeled as identically and independently distributed binomial variables
\begin{equation*}
\text{pr}(X_{i(T)} = x_i; T, N, n) = \binom{T}{x_i} \left(\frac{n}{N}\right)^{x_i} \left(1 - \frac{n}{N}\right)^{T - x_i}
\end{equation*}
However, since we only observe $X_{i(T)} > 0$, the above distribution must be truncated at zero. For convenience purposes, it is also useful to partition $\mathcal{P}$ into $D = \bigcup_{t=1}^T S_t = \{i \in \mathcal{P}: X_{i(T)} > 0\}$ and $U = \mathcal{P} \setminus D = \{i \in \mathcal{P}: X_{i(T)} = 0\}$. Accordingly, population size can be expressed as a sum of cardinalities of $D$ and $U$, i.e. $N = N_D + N_U$, where the first term is known and the latter term is to be estimated.  The resulting likelihood of the data is
\begin{equation*}
\mathcal{L}(N_U) = \prod_{i \in D} \frac{\text{pr}(X_{i(T)} = x_i; T, N, n)}{1 - \text{pr}(X_{i(T)} = 0; T, N, n)} = \prod_{i \in D} \binom{T}{x_i} \frac{n^{x_i} (N_D + N_U - n)^{T-x_i}}{(N_D + N_U)^{T} - (N_D + N_U - n)^T} \\
\end{equation*}
Setting $T = 2$ and recognizing that $\sum_{i \in D} x_i = nT$ yields the following maximum likelihood estimator of $N_U$
\begin{equation} \label{eq:0}
\widehat{N}_U = \frac{(n - N_D)^2}{2n - N_D}
\end{equation}
Thus, the population size is estimated by
\begin{equation*}
\widehat{N} = N_D + \widehat{N}_U = \frac{n^2}{2n - N_D}
\end{equation*}
The above expression can be rewritten in a more familiar manner, if we introduce the set of recaptured individuals $R = S_1 \cap S_2 = \{i \in \mathcal{P}: X_{i(2)} = 2\}$. The size of $D = S_1 \cup S_2$ is then the sum of sample sizes at each draw minus the size of $R$
\begin{equation} \label{eq:1}
\widehat{N} = \frac{n^2}{2n - 2n + N_R} = \frac{n^2}{N_R}
\end{equation}
Equation \eqref{eq:1} is the special case of the Lincoln-Peterson estimator with fixed $n$ for each sample draw \citep{Pollock:1990}.

Assume now that $\pi_i$ varies across population units meaning that \eqref{eq:1} is no longer applicable. Let us consider the problem from the Bayesian perspective and assume a beta prior for inclusion probabilities with hyperparameters $\alpha > 0$ and $\beta > 0$
\begin{equation*}
f_{\pi_i}(\upsilon; \alpha, \beta) = \frac{\upsilon^{\alpha - 1}(1 - \upsilon)^{\beta - 1}}{\mathrm{B}(\alpha, \beta)}, \quad \mathrm{B}(\alpha, \beta) = \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha + \beta)};
\end{equation*}
where $\Gamma(x)$ is the gamma function. From Bayes' theorem, we have
\begin{equation*}
f_{\pi_i \mid X_{i(T)}} (\upsilon \mid x_i) = \frac{\text{pr}(X_{i(T)} = x_i \mid \upsilon; T) f_{\pi_i}(\upsilon; \alpha, \beta)}{\text{pr}(X_{i(T)} = x_i)}
\end{equation*}
The marginal likelihood can be found in a straightforward manner
\begin{equation*}
\text{pr}(X_{i(T)} = x_i) = \int_0^1 \text{pr}(X_{i(T)} = x_i | \upsilon; T)f_{\pi_i}(\upsilon; \alpha, \beta)d\upsilon = \binom{T}{x_i} \frac{\mathrm{B}(\alpha + x_i, \beta + T - x_i)}{\mathrm{B}(\alpha, \beta)}
\end{equation*}
Conditioning on $X_{i(T)} > 0$ results in the following truncated distribution
\begin{equation*}
\text{pr}(X_{i(T)} = x_i \mid X_{i(T)} > 0; \alpha, \beta, T) = \frac{\text{pr}(X_{i(T)} = x_i)}{1 - \text{pr}(X_{i(T)} = 0)} = \binom{T}{x_i} \frac{\mathrm{B}(\alpha + x_i, \beta + T - x_i)}{\mathrm{B}(\alpha, \beta) - \mathrm{B}(\alpha, \beta + T)}
\end{equation*}
Following the empirical Bayes approach, we estimate hyperparameters $\alpha$ and $\beta$ by maximizing the marginal likelihood.
\begin{align*}
    \mathcal{L}(\alpha, \beta) &= \prod_{i \in D} \binom{T}{x_i} \frac{\mathrm{B}(\alpha + x_i, \beta + T - x_i)}{\mathrm{B}(\alpha, \beta) - \mathrm{B}(\alpha, \beta + T)} \\
    &= \prod_{i \in D} \binom{T}{x_i} \frac{\Gamma(\alpha + x_i)\Gamma(\beta + T - x_i)\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)\Gamma(\alpha + \beta + T) - \Gamma(\alpha)\Gamma(\beta + T)\Gamma(\alpha + \beta)} \\
\end{align*}
Using the recursive relation of the gamma function $\Gamma(x) = (x - 1)\Gamma(x - 1)$ and the fact that $T$ and $x_i$ are natural numbers, we can rewrite the marginal likelihood as
\begin{align} \label{eq:2}
    \mathcal{L}(\alpha, \beta) &= \prod_{i \in D} \binom{T}{x_i} \frac{\Gamma(\alpha)\Gamma(\beta)\Gamma(\alpha + \beta) \prod_{j=1}^{x_i} (\alpha + x_i - j)\prod_{j=1}^{T - x_i} (\beta + T - x_i - j)}{\Gamma(\alpha)\Gamma(\beta)\Gamma(\alpha + \beta)\{\prod_{j=1}^T(\alpha + \beta + T - j) - \prod_{j=1}^T (\beta + T - j)\}} \nonumber \\
    &= \prod_{i \in D} \binom{T}{x_i} \frac{\prod_{j=1}^{x_i} (\alpha + x_i - j)\prod_{j=1}^{T - x_i} (\beta + T - x_i - j)}{\prod_{j=1}^T(\alpha + \beta + T - j) - \prod_{j=1}^T (\beta + T - j)}
\end{align}
For inclusion probabilities $\pi_i$, it holds
\begin{equation} \label{eq:3}
E(\sum_{i \in \mathcal{P}} \pi_i) = N \frac{\alpha}{\alpha + \beta} = n
\end{equation}
The marginal likelihood \eqref{eq:2} is to be maximized subject to \eqref{eq:3}. Note that \eqref{eq:3} can also be viewed as centering our prior at $n / N$. In other words, the hyperparameters determine only the variance of the prior. From the constraint \eqref{eq:3}, we have $\beta = Nn^{-1}\alpha - \alpha$. Thus, the likelihood can be reparametrized in terms of $N_U$ and $\alpha$
\begin{equation} \label{eq:4}
    \mathcal{L}(\alpha, N_U) = \prod_{i \in D} \binom{T}{x_i} \frac{\prod_{j=1}^{x_i} (\alpha + x_i - j)\prod_{j=1}^{T - x_i} (Nn^{-1} \alpha - \alpha + T - x_i - j)}{\prod_{j=1}^T(Nn^{-1} \alpha + T - j) - \prod_{j=1}^T (Nn^{-1} \alpha - \alpha + T - j)}
\end{equation}
where $N = N_D + N_U$.

The maximum likelihood estimates of $N_U$ and $\alpha$ have no closed-form expressions and must be found numerically.

Setting $T = 2$ and $\alpha \to \infty$ will yield the estimator of $N_U$ equivalent to \eqref{eq:0}, as the variance of the prior tends to 0 resulting in a degenerate distribution centered at $n / N$. This can be shown more formally by taking the limit of \eqref{eq:4} with respect to $\alpha \to \infty$.

First, partition $D$ into the set of recaptured individuals $R = S_1 \cap S_2 = \{i \in \mathcal{P}: X_{i(2)} = 2\}$ and the set of individuals captured only once $C = \{i \in \mathcal{P}: X_{i(2)} = 1\}$ with sizes $N_R$ and $N_C$ respectively. Then, the likelihood simplifies to
\begin{align*}
    \mathcal{L}(\alpha, N_U) &= \prod_{i \in R \cup C} \binom{2}{x_i} \frac{\prod_{j=1}^{x_i} (\alpha + x_i - j)\prod_{j=1}^{2 - x_i} (Nn^{-1} \alpha - \alpha + 2 - x_i - j)}{\prod_{j=1}^2(Nn^{-1} \alpha + 2 - j) - \prod_{j=1}^2 (Nn^{-1} \alpha - \alpha + 2 - j)} \\
    &= \frac{\prod_{i \in C} 2\alpha(Nn^{-1}\alpha - \alpha + 1)\prod_{i \in R} \alpha (\alpha + 1)}{(2Nn^{-1}\alpha^2 - \alpha^2 + \alpha)^{N_C + N_R}} \\
    &= \frac{\alpha^{N_C + N_R}(\alpha + 1)^{N_R}\{2(Nn^{-1}\alpha - \alpha + 2)\}^{N_C}}{(2Nn^{-1}\alpha^2 - \alpha^2 + \alpha)^{N_C + N_R}} \\
    &= \left\{\frac{\alpha + 1}{\alpha(2Nn^{-1} - 1) + 1}\right\}^{N_R} \left\{\frac{2\alpha(Nn^{-1} - 1) + 4}{\alpha(2Nn^{-1} - 1) + 1}\right\}^{N_C}
\end{align*}
The limit of this expression as $\alpha \to \infty$ is
\begin{align} \label{eq:5}
&\lim_{\alpha \to \infty} \left[\left\{\frac{\alpha + 1}{\alpha(2Nn^{-1} - 1) + 1}\right\}^{N_R} \left\{\frac{2\alpha(Nn^{-1} - 1) + 4}{\alpha(2Nn^{-1} - 1) + 1}\right\}^{N_C}\right] \nonumber \\
&= \lim_{\alpha \to \infty} \left[\left\{\frac{1 + \alpha^{-1}}{2Nn^{-1} - 1 + \alpha^{-1}}\right\}^{N_R} \left\{\frac{2(Nn^{-1} - 1) + 4\alpha^{-1}}{2Nn^{-1} - 1 + \alpha^{-1}}\right\}^{N_C}\right] \nonumber \\
&= \frac{2^{N_C}(Nn^{-1} - 1)^{N_C}}{(2Nn^{-1} - 1)^{N_D}}
\end{align}
since $N_R + N_C = N_D$. By taking the logarithm of \eqref{eq:5} and maximizing with respect to $N$ we have
\begin{equation*}
\widehat{N} = \frac{n(2N_D - N_C)}{2(N_C - N_D)}
\end{equation*}
As $N_C = N_D - N_R$ and $N_D = 2n - N_R$, we arrive to
\begin{equation*}
\widehat{N} = \frac{n^2}{N_R}
\end{equation*}
which is the Lincoln-Peterson estimator for fixed $n$ as in \eqref{eq:1}. From it, the estimator \eqref{eq:0} follows.

\section{Discussion}
\begin{thebibliography}{1}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
